---
title: "NHANES-2015-2016"
author: "Elena Colicino, Michele Salerno"
date: "2025-04"
output: html_document
---

\
```{r results='asis', echo=FALSE}
cat("
## Download NHANES Datasets
1. [Download Demographic dataset](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/DEMO_I.xpt)
2. [Download Phthalates dataset](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/PHTHTE_I.xpt)
3. [Download Creatine/Albumin dataset](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/ALB_CR_I.xpt)
4. [Download Antibody dataset](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/SSPT_I.xpt)
")
```
\

# Dataset import

\
```{r}
#packages
library(haven)
library(dplyr)
library(tidyr)
library(DT)
library(htmltools)
library(skimr)
library(Hmisc)
library(knitr)
library(kableExtra)
library(ggplot2)
library(htmlwidgets)
library(plotly)

library(qgcomp)
library(bkmr)
library(FactoMineR)
library(factoextra)

library(data.table)
library(infinitefactor)
library(reshape2)

# ---- Define your data folder path ----
path.data = 'your/path/to/data/folder/'
#Demographic dataset
data_demo = read_xpt(paste0(path.data,"DEMO_I.XPT"))

#focus on children <18y
demo_under_18 = data_demo %>%
  filter(RIDAGEYR < 18) %>%
  select(SEQN, RIDAGEYR, RIAGENDR, RIDRETH1, RIDRETH3, INDFMPIR)
datatable(
  demo_under_18,
  extensions = 'Buttons',  
  options = list(
    dom = 'Bfrtip',  
    buttons = c('copy', 'csv', 'excel', 'print'),  
    pageLength = 10  
  ),
  rownames = FALSE
)

#Phthalates Dataset
data_phthalates = read_xpt(paste0(path.data,"PHTHTE_I.XPT"))

#Creatine/Albumin Dataset
data_creatine = read_xpt(paste0(path.data,"ALB_CR_I.XPT"))

#Antibody Dataset - tetanus and pertussis
data_antibody = read_xpt(paste0(path.data,"SSPT_I.XPT"))

#Merged dataset
merged_data = demo_under_18 %>%
  inner_join(data_phthalates, by = "SEQN") %>%
  inner_join(data_creatine, by = "SEQN") %>%
  inner_join(data_antibody, by = "SEQN") %>%
  drop_na()

#Create a single categorical variable for race/ethnicity
merged_data <- merged_data %>%
  mutate(
    RIDRETH3_recat = case_when(
      RIDRETH3 == 3 ~ "White",
      RIDRETH3 == 4 ~ "Black",
      RIDRETH3 %in% c(1, 2) ~ "Hispanic",
      RIDRETH3 %in% c(6, 7) ~ "Others"
    ) %>%
    factor(levels = c("White", "Black", "Hispanic", "Others"))
  ) %>%
  filter(RIDRETH3_recat != "Others") %>%
  mutate(RIDRETH3_recat = droplevels(RIDRETH3_recat))
```
\

# Descriptive Analysis

## Demographic variables

\
```{r}
# Age and Income
skim_without_charts(merged_data %>% select(RIDAGEYR, INDFMPIR))

# Age as categorical
merged_data <- merged_data %>%
  mutate(age_category = case_when(
    RIDAGEYR >= 6 & RIDAGEYR <= 9 ~ "6-9",
    RIDAGEYR >= 10 & RIDAGEYR <= 13 ~ "10-13",
    RIDAGEYR >= 14 & RIDAGEYR <= 17 ~ "14-17",
    TRUE ~ NA_character_
  ))

table(merged_data$age_category)

# Gender (1 = boys, 2 = girls)
table(merged_data$RIAGENDR)

# Ethnicity
table(merged_data$RIDRETH3_recat)                   
```
\

## Descriptive statistics antibodies for tetanus and pertussis 

\
```{r}
skim_without_charts(merged_data[, c("SSTT", "SSPT")])
```
\

# Phthalates adjusted for urine dilution: creatinine levels

\
```{r}
selected_vars = merged_data[, c(
                    "URXCNP","URXCOP","URXECP","URXHIBP","URXMBP","URXMC1",
                    "URXMCOH","URXMEP","URXMHBP","URXMHH","URXMHNC","URXMHP",
                    "URXMIB","URXMNP","URXMOH","URXMZP")]
adjusted_vars = selected_vars / merged_data$URXUCR
colnames(adjusted_vars) = paste0(colnames(adjusted_vars), "_adjusted")
merged_data1 = cbind(merged_data[,c("SEQN","RIDAGEYR","RIAGENDR","RIDRETH1","RIDRETH3","INDFMPIR", "RIDRETH3_recat",
                                    "SSPT", "SSTT")], adjusted_vars) 

skim_without_charts(merged_data1[, colnames(adjusted_vars)]) 
```
\

# Box-plot

## Box-plot antibody

\
```{r}
antibodies <- c("SSPT", "SSTT")
log_data_antibodies <- merged_data1 %>%
  select(all_of(antibodies)) %>%
  mutate(across(everything(), ~ log2(.))) %>%
  pivot_longer(cols = everything(), 
               names_to = "Antibody", values_to = "Log_Value") %>%
  mutate(Antibody = gsub("_adjusted", "", Antibody))

ggplot(log_data_antibodies, aes(x = Antibody, y = Log_Value)) +
  geom_boxplot(fill = NA) +  
  labs(x = "Antibodies", 
       y = "Log2(Concentration) (ng/mL)", 
       title = "Boxplot of Antibodies (Log2 Transformed)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 10),
    axis.text.y = element_text(size = 8)
  )
```
\

\
```{r}
merged_data1$Female = factor(ifelse(merged_data1$RIAGENDR == 2, 1, 0), levels=c(0,1))
```
\

\
```{r}
# phthalates log2 transformed
merged_data1 = merged_data1 %>%
  mutate(across(all_of(colnames(adjusted_vars)), 
                ~ log2(.), 
                .names = "{.col}_log"))

# antibody log2 transformed
merged_data1 = merged_data1 %>%
  mutate(across(all_of(antibodies), 
                ~ log2(.), 
                .names = "{.col}_log"))

```
\

## Outliers check

\
```{r}
detect_outliers = function(x) {
  Q1 = quantile(x, 0.25, na.rm = TRUE)
  Q3 = quantile(x, 0.75, na.rm = TRUE)
  IQR = Q3 - Q1
  lower_limit = Q1 - 3 * IQR
  upper_limit = Q3 + 3 * IQR
  outliers = which(x < lower_limit | x > upper_limit)
  return(outliers)
}

all_outliers = c()

for (phthalate in colnames(adjusted_vars)) {
  phthalate_log = paste0(phthalate, "_log")
  outliers_phthalate = detect_outliers(merged_data1[[phthalate_log]])
  all_outliers = c(all_outliers, outliers_phthalate)
  cat("Outliers in", phthalate_log, ": ", outliers_phthalate, "\n")
}

for (antibody in antibodies) {
  antibody_log = paste0(antibody, "_log")
  outliers_antibody = detect_outliers(merged_data1[[antibody_log]])
  all_outliers = c(all_outliers, outliers_antibody)
  cat("Outliers in", antibody_log, ": ", outliers_antibody, "\n")
}

all_outliers = unique(all_outliers)
merged_data1 = merged_data1[-all_outliers, ]
```
\

# LINEAR ANALYSIS

\
```{r}
# SSPT (Pertussis toxin antibody)

antibody_vars <- c("SSPT")

coefficients_summary <- data.frame(
  Number = integer(),
  Antibody = character(),
  Phthalate = character(),
  Coefficient = numeric(),
  `Std. Error` = numeric(),
  `p-value` = numeric(),
  `Lower_Confidence` = numeric(),
  `Upper_Confidence` = numeric(),
  p_adjusted_fdr = numeric(),
  stringsAsFactors = FALSE
)

for (antibody in antibody_vars) {
  
  p_values <- c() 

  for (i in 1:length(colnames(adjusted_vars))) {
    phthalate <- colnames(adjusted_vars)[i]

    phthalate_log <- paste0(phthalate, "_log")

    formula <- as.formula(
      paste0(
        antibody, "_log ~ ", phthalate_log,
        " + RIDAGEYR + Female + RIDRETH3_recat + INDFMPIR"
      )
    )

    model <- lm(formula, data = merged_data1)

    phthalate_coefficient <- summary(model)$coefficients[phthalate_log, 1]
    phthalate_se <- summary(model)$coefficients[phthalate_log, 2]
    phthalate_p <- summary(model)$coefficients[phthalate_log, 4]
    conf_interval <- confint(model, level = 0.95)[phthalate_log, ]

    p_values <- c(p_values, phthalate_p) 

    row_to_add <- data.frame(
      Number = i,
      Antibody = antibody,
      Phthalate = phthalate,
      Coefficient = phthalate_coefficient,
      `Std. Error` = phthalate_se,
      `p-value` = phthalate_p,
      `Lower_Confidence` = conf_interval[1],
      `Upper_Confidence` = conf_interval[2],
      p_adjusted_fdr = NA, 
      stringsAsFactors = FALSE
    )

    coefficients_summary <- rbind(coefficients_summary, row_to_add)
  }

  adjusted_p_values <- p.adjust(p_values, method = "fdr")
  coefficients_summary$p_adjusted_fdr[
    coefficients_summary$Antibody == antibody
  ] <- adjusted_p_values
}

datatable(
  coefficients_summary,
  extensions = 'Buttons',  
  options = list(
    dom = 'Bfrtip',  
    buttons = c('copy', 'csv', 'excel', 'print'),  
    pageLength = 10
  ),
  rownames = FALSE
)


coefficients_summary$Coefficient = ((2^(coefficients_summary$Coefficient))-1)*100
coefficients_summary$Lower_Confidence = ((2^(coefficients_summary$Lower_Confidence))-1)*100
coefficients_summary$Upper_Confidence = ((2^(coefficients_summary$Upper_Confidence))-1)*100


datatable(
  coefficients_summary,
  extensions = 'Buttons',  
  options = list(
    dom = 'Bfrtip',  
    buttons = c('copy', 'csv', 'excel', 'print'),
    pageLength = 10
  ),
  rownames = FALSE
)
```
\

## Forest plot

\
```{r}
for (antibody in antibody_vars) {
  data_subset = coefficients_summary %>% filter(Antibody == antibody)
  
  if (nrow(data_subset) > 0) {
    plot = ggplot(data_subset, aes(x = reorder(Phthalate, Coefficient), y = Coefficient)) +
      geom_point(size = 3, color = "#3CB371") +
      geom_errorbar(aes(ymin = Lower_Confidence, ymax = Upper_Confidence), 
                    width = 0.2, color = "#3CB371") +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +  
      coord_flip() +
      theme_minimal() +
      labs(
        title = paste("Forest Plot for Antibody:", antibody),
        x = "Phthalates",
        y = "Coefficient (95% CI)"
      )
    
    print(plot)
  } 
}
```

\

# QUANTILE G-COMPUTATION APPROACH

\
```{r}
Xnm <- c("URXMEP_adjusted_log", "URXMBP_adjusted_log", "URXMIB_adjusted_log",
         "URXHIBP_adjusted_log", "URXMC1_adjusted_log", "URXMHBP_adjusted_log", 
         "URXMZP_adjusted_log", "URXMHH_adjusted_log", "URXMOH_adjusted_log", 
         "URXECP_adjusted_log", "URXMCOH_adjusted_log", "URXCNP_adjusted_log",
         "URXCOP_adjusted_log", "URXMHNC_adjusted_log", "URXMNP_adjusted_log",
         "URXMHP_adjusted_log"
)
covars = c("RIDAGEYR", "Female", "RIDRETH3_recat", "INDFMPIR")

### Quadratic
sspt_quadr = qgcomp.glm.boot(
  SSPT_log ~ RIDAGEYR + Female + RIDRETH3_recat + INDFMPIR + . + .^2,
  expnms = Xnm,
  data = merged_data1,
  family = gaussian(),
  q = 10,
  B = 1000,
  degree = 2
)

summary(sspt_quadr)
plot(sspt_quadr)

### Quadratic
sspt_quadr = qgcomp.glm.noboot(SSPT_log ~ RIDAGEYR + Female + RIDRETH3_recat + INDFMPIR + . + .^2, 
       expnms = Xnm, 
       data = merged_data1[, c(Xnm, 'SSPT_log', 'RIDAGEYR', 'Female', 'RIDRETH3_recat', 'INDFMPIR')], 
       family = gaussian(), 
       q = 10)
summary(sspt_quadr)
plot(sspt_quadr)

```
\

# BKMR

# SSPT

\
```{r}
y = merged_data1$SSPT_log

Z = scale(merged_data1 %>%
            select(URXMEP_adjusted_log, URXMBP_adjusted_log, URXMIB_adjusted_log, 
                  URXHIBP_adjusted_log, URXMC1_adjusted_log, URXMHBP_adjusted_log, 
                  URXMZP_adjusted_log, URXMHH_adjusted_log, URXMOH_adjusted_log, 
                  URXECP_adjusted_log, URXMCOH_adjusted_log, URXCNP_adjusted_log, 
                  URXCOP_adjusted_log, URXMHNC_adjusted_log, URXMNP_adjusted_log, URXMHP_adjusted_log))

X = merged_data1 %>%
  mutate(
    RIDAGEYR = scale(RIDAGEYR),  
    INDFMPIR = scale(as.numeric(INDFMPIR)),  
    Female = as.numeric(as.character(Female)),
    RIDRETH3_recat = scale(as.numeric(RIDRETH3_recat))
  ) %>%
  select(RIDAGEYR, Female, INDFMPIR, RIDRETH3_recat)
```
\

\
```{r}
# model
set.seed(123)
fitkm = kmbayes(y = y, Z = Z, X = X, iter = 10000, verbose = TRUE, varsel = TRUE)
```
\

## Trace plot

Let's print trace plots to check convergence:

\
```{r}
# Trace plot for coefficients of the covariates
TracePlot(fit = fitkm, par = "beta")
# Trace plot for residual variance
TracePlot(fit = fitkm, par = "sigsq.eps")
# Trace plot for the relevance of exposures
TracePlot(fit = fitkm, par = "r", comp = 1)
```
\

## PIP

Estimated posterior inclusion probabilities:

\
```{r}
ExtractPIPs(fitkm)
```
\

**Interpretation of PIPs**:

- If the PIP of a variable is 1.0000, it means that that variable was included in all iterations of posterior sampling and thus was selected as relevant.

- If the PIP is low (e.g., 0.10), the variable was included in only a few iterations and therefore has less influence on the model.

# Summarize model output

\
```{r}
pred.resp.univar <- PredictorResponseUnivar(fit = fitkm)
```
\

\
```{r}
ggplot(pred.resp.univar, aes(z, est, ymin = est - 1.96*se, ymax = est + 1.96*se)) +  
  geom_smooth(stat = "identity") +  
  facet_wrap(~ variable, labeller = labeller(variable = function(x) gsub("_adjusted_log", "", x))) +  
  ylab("h(z)")
```
\

## Summary statistics of the predictor-response function

Overall risk (overall risk) associated with exposures (Z) in the BKMR model.

\
```{r}
X_clean = data.frame(lapply(X, function(col) {
  attributes(col) <- NULL
  return(col)
}))

risks.overall <- OverallRiskSummaries(
  fit = fitkm, 
  y = y, 
  Z = Z, 
  X = as.matrix(X_clean),
  qs = seq(0.25, 0.75, by = 0.05), 
  q.fixed = 0.5, 
  method = "exact"
)
print(risks.overall)
```
\

\
```{r}
ggplot(risks.overall, aes(quantile, est, ymin = est - 1.96*sd, ymax = est + 1.96*sd)) + 
    geom_pointrange()
```
\

### Individual predictors

Summary analysis of the risks associated with individual predictors (Z) in the BKMR model, calculating the risk associated with a single predictor at different levels (percentiles) of the other predictors. 

\
```{r}
X_clean = data.frame(lapply(X, function(col) {
  attributes(col) <- NULL
  return(col)
}))

risks.singvar = SingVarRiskSummaries(
  fit = fitkm, 
  y = y, 
  Z = Z, 
  X = as.matrix(X_clean), 
  qs.diff = c(0.25, 0.75), 
  q.fixed = c(0.25, 0.50, 0.75), 
  method = "exact"
)
print(risks.singvar)
```
\

\
```{r}
ggplot(risks.singvar, aes(variable, est, ymin = est - 1.96*sd, 
                          ymax = est + 1.96*sd, col = q.fixed)) + 
    geom_pointrange(position = position_dodge(width = 0.75)) + 
  coord_flip()
```
\

### Interaction effects

Estimating and visualizing the interaction effects between a single predictor and all other predictors in the BKMR model.

\
```{r}
X_clean = data.frame(lapply(X, function(col) {
  attributes(col) <- NULL
  return(col)
}))

risks.int = SingVarIntSummaries(
  fit = fitkm, 
  y = y, 
  Z = Z, 
  X = as.matrix(X_clean), 
  qs.diff = c(0.25, 0.75), 
  qs.fixed = c(0.25, 0.75), 
  method = "exact" 
)
print(risks.int)
```
\

# Bayesian Infinite Factor Models

\
```{r}
merged_data1 = merged_data1 %>%
  mutate(
    Ethnicity_code = case_when(
      RIDRETH3_recat == "White" ~ 1,
      RIDRETH3_recat == "Black" ~ 2,
      RIDRETH3_recat == "Hispanic" ~ 3,
    )
  ) 
```
\

## PCA

\
```{r}
phthalates_scaled = scale(merged_data1 %>% select(all_of(colnames(adjusted_vars))))

# PCA
pca_results = PCA(phthalates_scaled, graph = FALSE)

eig_vals <- pca_results$eig
var_explained <- eig_vals[, 2]

cum_var_explained <- cumsum(var_explained)

# Scree plot -> elbow point
fviz_eig(pca_results, addlabels = TRUE, ylim = c(0, 50))

# To find the number of components that explain at least 50% of the variance
n_components <- which(cum_var_explained >= 50)[1] 
cat("Number of factors to consider:", n_components, "\n")

summary(pca_results)

# Variance Explained and Cumulative Variance
data.frame(Component = 1:length(var_explained),
           Variance_Explained = var_explained,
           Cumulative_Variance = cum_var_explained)
```
\

\[
\boldsymbol{\beta}_X = \mathbf{\Lambda} \cdot \boldsymbol{\omega}
\]

where:

- $\mathbf{\Lambda}$ is the **factor loading matrix** ($p \times k$), with:
  - $p$ = number of metabolites (observed variables),
  - $k$ = number of latent factors.

- $\boldsymbol{\omega}$ is the **regression coefficient vector** ($k \times 1$) obtained from the regression $y \sim \text{latent factors}$.

Thus, the estimated effect for each metabolite $i$ is:

\[
\beta_{X_i} = \sum_{j=1}^{k} \lambda_{ij} \cdot \omega_j
\]

where:
- $\lambda_{ij}$ represents the loading of metabolite $i$ on latent factor $j$,
- $\omega_j$ is the estimated regression coefficient for factor $j$.

## SSPT

\
```{r}
# Creating the X matrix
setDT(merged_data1)

phthalates = c(
  "URXMEP_adjusted_log", "URXMBP_adjusted_log", "URXMIB_adjusted_log", 
  "URXHIBP_adjusted_log", "URXMC1_adjusted_log", "URXMHBP_adjusted_log", 
  "URXMZP_adjusted_log", "URXMHH_adjusted_log", "URXMOH_adjusted_log", 
  "URXECP_adjusted_log", "URXMCOH_adjusted_log", "URXCNP_adjusted_log", 
  "URXCOP_adjusted_log", "URXMHNC_adjusted_log", "URXMNP_adjusted_log", "URXMHP_adjusted_log"
)

merged_data1[, ':='(
  white = ifelse(Ethnicity_code == 1, 1, 0),
  black = ifelse(Ethnicity_code == 2, 1, 0),
  hispanic = ifelse(Ethnicity_code == 3, 1, 0)
)]

merged_data1$Female <- as.numeric(merged_data1$Female)

y = as.matrix(merged_data1$SSPT_log)

X = as.matrix(merged_data1[, .SD, .SDcols = c(phthalates, "RIDAGEYR", "Female", "INDFMPIR", "black", "hispanic")])

covariates = c("RIDAGEYR", "Female", "INDFMPIR", "black", "hispanic")

#print(X)
```
\

\
```{r}
set.seed(123)

nrun = 10000
burn = 3000
thin = 10
k = 2
intdl = interactionDL(y = y, X = X, nrun = nrun, burn = burn, thin = thin, k = k)
```
\

\
```{r}
# Rotate 
Rotout = jointRot(intdl$lambdaSamps, intdl$etaSamps)
mLambda = lmean(Rotout$lambda)
rownames(mLambda) = colnames(X)

dim(Rotout$lambda[[1]]) 

# Create an array with all the simulation 
aLambda = array(NA, dim = c(21, 2, 9)) 
for(i in 1:9){
  aLambda[,,i] = Rotout$lambda[[i]][, 1:2]
}

# Calculate quantile for each component in the factor
aLambdaq1 = apply(aLambda, c(1,2), quantile, c(0.025))
aLambdaq2 = apply(aLambda, c(1,2), quantile, c(0.975))

mLambda1 = as.data.table(mLambda)
setnames(mLambda1,names(mLambda1),paste0("factor_",1:2))
mLambda1[,names := colnames(X)]

aLambdaq1 = as.data.table(aLambdaq1)
setnames(aLambdaq1,names(aLambdaq1),paste0("factor_",1:2))
aLambdaq1[,names := colnames(X)]

aLambdaq2 = as.data.table(aLambdaq2)
setnames(aLambdaq2,names(aLambdaq2),paste0("factor_",1:2))
aLambdaq2[,names := colnames(X)]

mLambda1 = melt(mLambda1)
aLambdaq1 = melt(aLambdaq1)
aLambdaq2 = melt(aLambdaq2)

mLambda1 = as.data.table(mLambda1)
aLambdaq1 = as.data.table(aLambdaq1)
aLambdaq2 = as.data.table(aLambdaq2)

aLambdaq1 = aLambdaq1[,.(names, variable=as.character(variable), value)]
mLambda1 = mLambda1[,.(names, variable=as.character(variable), value)]
aLambdaq2 = aLambdaq2[,.(names, variable=as.character(variable), value)]

# Table of the loadings  
data = Reduce(function(x, y) merge(x, y, 
    by=c("names", "variable")), list(aLambdaq1,mLambda1,aLambdaq2))
data[,check := ifelse(sign(value.x) == sign(value),1,0)]
setnames(data,names(data),c("Metabolite","Factor","q025","mean","q975","sig"))
data[order(sig,decreasing = T)]
```
\

\
```{r}
# Filter significant metabolites only
significant_loadings = data[q025 > 0 | q975 < 0]
significant_loadings = significant_loadings[Metabolite %in% phthalates]

# Sort by value
significant_loadings = significant_loadings[order(Factor, -abs(mean))]
significant_loadings
```
\

\
```{r}
# Heatmap
heatmap_data = significant_loadings[, .(Metabolite, Factor, mean)]
heatmap_matrix = dcast(heatmap_data, Metabolite ~ Factor, value.var = "mean")

ggplot(melt(heatmap_matrix, id.vars = "Metabolite"), aes(x = variable, y = Metabolite, fill = value)) +  
  geom_tile(colour = "grey20") +  
  scale_fill_gradient2(low = "#3d52bf", high = "#33961b", mid = "white", midpoint = 0) +  
  theme_minimal() +  
  labs(x = "Factor", y = "Metabolite", fill = "Loading") +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 8),
        legend.title = element_text(),
        plot.title = element_text(hjust = 0.5)) +  
  ggtitle("Heatmap of Significant Factor Loadings")
```
\

### Estimated Main Effects

\
```{r}
# Find Estimated Main Effects

# Convert list to 3D array: (iterations, individuals, factors)
factors_array = simplify2array(intdl$etaSamps)  # (387, 2, 9)

# Check dimension
#dim(factors_array)  

# Calculate the mean over the MCMC samples to obtain a matrix (387 individuals × 2 factors)
factors_matrix = apply(factors_array, c(1, 2), mean)

# Check matrix dimension's
#dim(factors_matrix) 

# Converts to dataframe for regression
factors_df = as.data.frame(factors_matrix)
colnames(factors_df) = paste0("Factor_", 1:ncol(factors_df))

# Regression of the variable of interest (y) on the latent factors (omega)
lm_results = lm(y ~ ., data = cbind(y = y, factors_df))

summary(lm_results)
```
\

\[
\boldsymbol{\beta}_X = \mathbf{\Lambda} \cdot \boldsymbol{\omega}
\]

where:

- $\mathbf{\Lambda}$ is the **factor loading matrix** ($p \times k$), with:
  - $p$ = number of metabolites (observed variables),
  - $k$ = number of latent factors.

- $\boldsymbol{\omega}$ is the **regression coefficient vector** ($k \times 1$) obtained from the regression $y \sim \text{latent factors}$.

Thus, the estimated effect for each metabolite $i$ is:

\[
\beta_{X_i} = \sum_{j=1}^{k} \lambda_{ij} \cdot \omega_j
\]

where:
- $\lambda_{ij}$ represents the loading of metabolite $i$ on latent factor $j$,
- $\omega_j$ is the estimated regression coefficient for factor $j$.

\
```{r}
omega_estimates = coef(lm_results)[-1]
# Make sure omega_estimates is a column matrix (2 x 1)
omega_estimates = matrix(omega_estimates, ncol = 1)

# Estimated Main Effects
beta_X_estimates = mLambda %*% omega_estimates # mLambda = Factor Loadings
beta_X_estimates
```
\

$\boldsymbol{\beta}_X$ represents the vector of **estimated main effects**, i.e., the associations between each **metabolite** and the outcome variable $y$. Thus, $\boldsymbol{\beta}_X$ allows us to **project the effects of latent factors onto the observed metabolites**, determining their estimated influence on the outcome variable.

\
```{r}
# N MCMC iterations saved
n_iter = length(Rotout$lambda)

beta_X_samples = matrix(NA, nrow = n_iter, ncol = nrow(mLambda))

# Loop MCMC iterations
for (i in 1:n_iter) {
  lambda_sample = Rotout$lambda[[i]][, 1:2]  
  beta_X_samples[i, ] = lambda_sample %*% omega_estimates  
}

# Confidence intervals
beta_X_mean = apply(beta_X_samples, 2, mean)      # mean
beta_X_q025 = apply(beta_X_samples, 2, quantile, probs = 0.025)  # q25
beta_X_q975 = apply(beta_X_samples, 2, quantile, probs = 0.975)  # q75

beta_X_results = data.frame(
  Metabolite = rownames(mLambda),
  Beta = beta_X_mean,
  IC_lower = beta_X_q025,
  IC_upper = beta_X_q975
)

print(beta_X_results)
```
\

\
```{r}
ggplot(beta_X_results, aes(x = Metabolite, y = Beta)) +
  geom_point(color = "red", size = 3) +  
  geom_errorbar(aes(ymin = IC_lower, ymax = IC_upper), width = 0.2, color = "red") +  
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey", linewidth = 1) +  
  theme_minimal() +
  labs(
    x = "",
    y = "Estimated Main Effects",
    title = "Estimated Main Effects - SSPT"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )
```
\

\
```{r}
# Quadratic term
inter_m_mat = apply(intdl$omegaSamps, c(1,2), mean)
inter_q1_mat = apply(intdl$omegaSamps, c(1,2), quantile, c(0.025))
inter_q2_mat = apply(intdl$omegaSamps, c(1,2), quantile, c(0.975))

inter_m_mat = as.data.table(inter_m_mat)
inter_m_mat[,names := colnames(X)]
setnames(inter_m_mat,names(inter_m_mat),c(colnames(X),"names"))
inter_m_mat = melt(inter_m_mat)

inter_q1_mat = as.data.table(inter_q1_mat)
inter_q1_mat[,names := colnames(X)]
setnames(inter_q1_mat,names(inter_q1_mat),c(colnames(X),"names"))
inter_q1_mat = melt(inter_q1_mat)

inter_q2_mat = as.data.table(inter_q2_mat)
inter_q2_mat[,names := colnames(X)]
setnames(inter_q2_mat,names(inter_q2_mat),c(colnames(X),"names"))
inter_q2_mat = melt(inter_q2_mat)
```
\

\
```{r}
inter_m_mat = as.data.table(inter_m_mat)
inter_q1_mat = as.data.table(inter_q1_mat)
inter_q2_mat = as.data.table(inter_q2_mat)

data_inter = Reduce(function(x, y) merge(x, y, 
    by=c("names", "variable")), list(inter_q1_mat[,.(names, variable=as.character(variable), value)],
                                     inter_m_mat[,.(names, variable=as.character(variable), value)],
                                     inter_q2_mat[,.(names, variable=as.character(variable), value)]))
data_inter[,check := ifelse(sign(value.x) == sign(value),1,0)]

data_inter_tmp = data_inter[names %in% phthalates & variable %in% phthalates]
data_inter_tmp[, value_plot := ifelse(check == 0, NA, value.y)]
setkey(data_inter_tmp,names,variable)
```
\

\
```{r}
ggplot(data_inter_tmp, aes(names, variable, fill = value_plot)) + 
  geom_tile() +
  scale_y_discrete(limits = unique(data_inter_tmp$names)) +
  scale_x_discrete(limits = rev(unique(data_inter_tmp$names))) +
  scale_fill_gradient2(low = "#f4d219", mid = "#99CCFF", high = "#336699") +
  ggtitle("Interaction Metabolites") +
  labs(x = "", y = "") +
  theme(plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),
        panel.grid = element_blank(),
        axis.text.x = element_text(size = 6, angle = 90, hjust = 0.95, vjust = 0.2),
        axis.text.y = element_text(size = 6))
```
\

### Interactions Phthalates-Covariates

\
```{r}
data_inter_tmp = data_inter[names %in% phthalates & variable %in% c("RIDAGEYR", "Female", "INDFMPIR", "black", "hispanic", "chol", "creatinine")]

#Only significant interactions (those that do not include 0 in the 99% CI)
data_inter_tmp[, value_plot := ifelse(check == 0, NA, value.y)]

ggplot(data_inter_tmp, aes(x = variable, y = names, fill = value_plot)) +  
  geom_tile(color = "grey20") +  
  scale_fill_gradient2(low = "#3d52bf", mid = "white", high = "#33961b") +  
  labs(x = "Covariate", y = "Phthalate", fill = "Interaction Coefficient") +  
  theme_minimal() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        plot.title = element_text(hjust = 0.5)) +  
  ggtitle("Significant Interactions Phthalates-Covariates")
```
\
